Channel Shaping Using Beyond Diagonal Reconfigurable Intelligent Surfaces

### 2024-05-14 main.tex
:triangular_flag_on_post: Bruno 240514

\usepackage[table]{xcolor}
\usepackage{adjustbox}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bookmark}
\usepackage{booktabs}
\usepackage[makeroom]{cancel}
\usepackage[american]{circuitikz}
\usepackage{cite}
\usepackage{fixmath}
\usepackage[acronym]{glossaries-extra}
\usepackage{hyperref}
\usepackage{import}
\usepackage{mathtools}
\usepackage{microtype}
\usepackage[short]{optidef}
\usepackage{pgfplots}
\usepackage{ragged2e}
\usepackage[subtle]{savetrees}
\usepackage{siunitx}
\usepackage{stfloats}
\usepackage[caption=false,font=footnotesize,subrefformat=parens,labelformat=parens]{subfig}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{multirow}
\usepackage{booktabs}
% \usepackage{cleveref}

@@

\begin{document}
\title{Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal Model to Beyond}
\author{
	\IEEEauthorblockN{
		Yang~Zhao,~\IEEEmembership{Member,~IEEE,}
		Hongyu~Li,~\IEEEmembership{Graduate Student Member,~IEEE,}\\
		Massimo~Franceschetti,~\IEEEmembership{Fellow,~IEEE,}
		and~Bruno~Clerckx,~\IEEEmembership{Fellow,~IEEE}
	}
	\thanks{
		Yang Zhao, Hongyu Li, and Bruno Clerckx are with the Department of Electrical and Electronic Engineering, Imperial College London, London SW7 2AZ, U.K. (e-mail: \{\href{mailto:yang.zhao18@imperial.ac.uk}{yang.zhao18}, \href{mailto:c.li21@imperial.ac.uk}{c.li21}, \href{mailto:b.clerckx@imperial.ac.uk}{b.clerckx}\}@imperial.ac.uk).

@@

\begin{section}{Introduction}
	Today we are witnessing a paradigm shift from connectivity to intelligence, where the wireless environment is no longer a chaotic medium but a conscious agent that can serve on demand.
	This is empowered by the recent advances in \gls{ris}, a programmable passive planar surface that recycles and redistributes ambient electromagnetic waves for improved wireless performance.
	A typical \gls{ris} consists of numerous low-power sub-wavelength non-resonant scattering elements, whose response can be engineered in real-time to manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019}.
	% It not only experiences negligible noise and supports full-duplex transmissions, but also features better flexibility than reflectarrays, lighter footprint than various relays, and greater scalability than conventional multi-antenna techniques.
	It enables low-noise, full-duplex operation, and also features better flexibility than reflectarrays, lighter footprint than various relays, and greater scalability than conventional \gls{mimo} systems.
	The most popular \gls{ris} research direction is \emph{joint passive and active beamforming} design with transceivers for a specific performance measure, which has attracted significant attention in wireless communication \cite{Wu2019,Guo2020,Liu2022}, sensing \cite{He2022,Luo2022,Hua2023}, and power transfer literature \cite{Wu2020a,Feng2022,Zhao2022}.
	While passive beamforming at \gls{ris} suffers severe attenuation from double fading, it usually offers a squared asymptotic behavior than active beamforming at transceivers (e.g., second-order array gain and fourth-order harvested power \cite{Zhao2022}).
	On the other hand, \gls{ris} can also be used for \emph{backscatter modulation} by periodically switching its reflection pattern within the channel coherence time.
	This creates a free-ride message stream (similar to index modulation \cite{Basar2017}) with dual benefits: integrating with the legacy transmitter for enhanced channel capacity \cite{Karasik2020,Basar2020,Ye2022} or serving as a dedicated source for low-power uplink communication \cite{Liang2020,Zhao2024,Yang2024}.
	Different from above directions, \emph{channel shaping} exploits the \gls{ris} as a stand-alone device to modify the inherent properties of the wireless.
	To name a few, it can compensate for the Doppler effect \cite{Basar2021}, transform frequency-selective channels into frequency-flat \cite{Arslan2022}, improve the spatial diversity for \gls{mimo} systems \cite{Ozdogan2020a}, and create artificial time diversity for orthogonal \cite{Yang2019} and non-orthogonal \cite{Chen2023} multiple accesses.
	Channel shaping also provides a ubiquitous benchmark for different wireless applications and helps to decouple the \gls{ris}-transceiver design.
	At a specific time and frequency, channel statistics can be roughly classified into two categories:
	% shaping metrics can be classified into two categories:
	\begin{itemize}
		\item \emph{Singular value:} Those metrics are in general closely related to the performance measures (e.g., achievable rate and harvested power \cite{Shen2021}) but sensitive to numerical perturbations of the channel matrix. The impact of \gls{ris} has been studied in terms of minimum singular value \cite{ElMossallamy2021}, effective rank \cite{ElMossallamy2021,Meng2023}, condition number \cite{Zheng2022,Huang2023}, and degree of freedom \cite{Bafghi2022,Zheng2023,Chae2023}.
		\item \emph{Power:} Those second-order statistics are less informative in \gls{mimo} but easier to analyze and optimize. The impact of \gls{ris} has been studied in terms of channel power gain \cite{Wu2019,Shen2020a,Nerini2023,Nerini2024,Santamaria2023} and leakage interference \cite{Santamaria2023a}.
	\end{itemize}

	Although those works offer initial glimpses into the channel shaping potential, one critical question remains unaddressed: \emph{To what extent can a passive \gls{ris} reshape the \gls{mimo} channel in terms of singular values?}

@@

	Most relevant works \cite{ElMossallamy2021,Meng2023,Zheng2022,Huang2023,Bafghi2022,Zheng2023,Chae2023,Wu2019,Santamaria2023a} follows a conventional path where each \gls{ris} element is tuned by a dedicated impedance and acts as an \emph{individual} scatterer \cite{Wu2020}.
	This simple architecture translates mathematically to a diagonal scattering matrix with unit-magnitude entries on the main diagonal and zeros elsewhere, applying merely a phase shift to the incoming signal.
	The concept was soon generalized to \gls{bd}-\gls{ris} with a group-connected architecture \cite{Shen2020a} that groups adjacent elements and connects them via passive reconfigurable components.\footnote{Those components can be either symmetric (e.g., capacitors and inductors) or asymmetric (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006}, resulting in symmetric and asymmetric scattering matrices, respectively.}
	This allows \emph{cooperative} scattering --- wave impinging on one element can propagate within the circuit and depart partially from any element in the same group.

@@

	The authors of \cite{Bartoli2023} investigated the rate-optimal joint beamforming design for a specific \gls{bd}-\gls{ris}-aided \gls{mimo} system with blocked direct link and unitary (a.k.a. fully-connected) \gls{ris}.

@@

	Furthermore, no previous work has attempted to characterize the singular value region of a \gls{mimo} channel manipulated by any type of \gls{ris}, and the \gls{bd}-\gls{ris}-aided \gls{mimo} rate maximization problem remains unsolved.

@@

	First, we interpret the channel shaping potential of \gls{bd}-\gls{ris} in terms of channel rearrangement and space alignment.
	Channel rearrangement refers to rearrange and recombine the forward and backward channel branches by their strength.
	Space alignment generalizes phase matching in \gls{siso} and \gls{miso} to the high-dimensional singular vector space in \gls{mimo}.
	The off-diagonal entries of \gls{bd}-\gls{ris} scattering matrix provides a higher design freedom and exploits spatial diversity more effectively than diagonal \gls{ris}.
	This is the first paper to study \gls{bd}-\gls{ris} in general \gls{mimo} systems.
	%  and discuss its shaping potential.

	Second, we propose an efficient \gls{bd}-\gls{ris} design framework based on geodesic\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} \gls{rcg}.
	% to numerically analyze the channel shaping capability of \gls{bd}-\gls{ris}, we exploit the Riemannian geometry of the Stiefel manifold and
	This method modified from \cite{Abrudan2008,Abrudan2009} exploits the Riemannian geometry of the Stiefel manifold, which not only converges faster than non-geodesic approach \cite{Absil2009,Pan2022d} but also works for general optimization problems of group-connected \gls{bd}-\gls{ris}.
	Specifically, group-wise multiplicative updates are performed along the geodesics to avoid retraction and facilitate step size selection.
	It greatly improves computational efficiency and stability.
	This is the first work to tailor an efficient and universal optimization framework for \gls{bd}-\gls{ris}.

	% pioneer
	Third, we adopt the proposed geodesic \gls{rcg} and provide a numerical answer to the singular value shaping question.
	% to answer the channel singular value shaping question via numerical optimization.
	% to redistribute the singular values of a point-to-point \gls{mimo} channel.
	The {Pareto frontiers} of channel singular values are characterized by solving a series of weighted sum maximization problems, where the weight associated with each singular value can be positive, zero, or negative.
	% optimizing their weighted sum, where the weights can be positive, zero, or negative.
	% This problem is solved by the proposed geodesic \gls{rcg} algorithm.
	The enclosed region generalizes most relevant metrics and provides an intuitive channel shaping benchmark.
	Results show that increasing \gls{bd}-\gls{ris} group size enlarges the singular value region, improving both dynamic range of and trade-off between singular values.
	% , where the dynamic range and trade-off between are both improved.
	% significantly enlarges the singular value region, improveing the dynamic range and trade-off in-between.
	% We also and derive some analytical singular value bounds for rank-deficient \gls{mimo} and unitary \gls{ris}.

	Fourth, we leverage matrix analysis tools and provide an analytical answer to the singular value shaping question.
	% revisit the channel singular value shaping problem from an analytical perspective via matrix analysis.
	% To explore the ultimate shaping limits, we derive individual and collective singular value bounds for rank-deficient forward/backward channel and blocked direct channel.
	% Asymptotic bounds shared by diagonal and \gls{bd}-\gls{ris} are derived in the former case, while exact bounds for unitary \gls{bd}-\gls{ris} are derived in the latter case.
	Asymptotic channel singular value bounds shared by diagonal and \gls{bd}-\gls{ris} are derived for rank-deficient forward/backward channel.
	On the other hand, attainable bounds in the presence of unitary \gls{bd}-\gls{ris} are derived for blocked direct channel.
	Results validate those bounds and emphasizes the shaping advantage of \gls{bd}-\gls{ris}.
	This is the first work to quantify the singular value redistribution capability from either numerical or analytical perspective.
	% Results validate those bounds and show that
	% fully-connected \gls{bd}-\gls{ris} without direct link.

@@

	% % respectively in rank-deficient channels and for fully-connected \gls{bd}-\gls{ris}
	% collective singular value bounds	for rank-deficient channels and unitary \gls{ris}, respectively.
	% 	Results validate those bounds and show
	% for rank-deficient channels and unitary \gls{ris}, respectively.
	% Results validate those bounds and show that increasing \gls{bd}-\gls{ris} group size significantly enlarges the singular value region, providing wider dynamic range and better trade-off.
	% dynamic range and trade-off of singular values are significantly improved by \gls{bd}-\gls{ris}, especially in large-scale \gls{mimo} systems.
	% This is the first work to comprehensively answer the channel shaping capability question from both numerical and analytical perspectives in the most general \gls{bd}-\gls{ris} setup.
	% from a Pareto perspective.

	Fifth, we tackle \gls{bd}-\gls{ris}-aided \gls{mimo} rate maximization problem by two beamforming solutions: a local-optimal \gls{ao} approach and a low-complexity approach inspired by channel shaping insights.
	The former updates active beamforming by eigenmode transmission and passive beamforming by geodesic \gls{rcg} until convergence.
	The latter suboptimally decouples the problem as channel power gain maximization at \gls{ris} and conventional \gls{mimo} transmission at transceiver, then solves both in closed form.
	% The problem is solved by a closed-form two-stage method.
	Results show that \gls{bd}-\gls{ris} provides higher power gain and achievable rate than diagonal \gls{ris}, and the rate gap between two approaches diminishes as the \gls{ris} evolves from diagonal to unitary.
	It suggests channel shaping offers a promising path to decouple joint \gls{ris}-transceiver designs.
	% Interestingly, the rate difference between the optimal and low-complexity designs diminishes as the \gls{ris} evolves from diagonal to unitary.
	% Those results emphasize the importance of using \gls{bd}-\gls{ris} in large-scale \gls{mimo} systems and suggest channel shaping offers a promising path to decouple joint \gls{ris}-transceiver designs.
	% a promising channel shaping path towards practical \gls{ris} and transceiver design.
	% that channel shaping offers a promising path towards simplified and practical \gls{bd}-
	% It suggests that channel shaping offers a promising path towards simplified and practical \gls{bd}-\gls{ris} (and transceiver) designs.
	% via \gls{bd}-\gls{ris} offers a promising path towards simplified and practical design.
	% offers a promising path towards simplified and practical desi
	% It suggests channel shaping helps to decouple the joint \gls{ris}-transceiver design and offers a promising low-complexity solution.
	% This is the first work to discuss the implications of channel shaping on joint active-passive beamforming optimization.

@@

\begin{section}{\glsfmtshort{bd}-\glsfmtshort{ris} Model}
	Consider a \gls{bd}-\gls{ris} aided point-to-point \gls{mimo} system with $N_\mathrm{T}$, $N_\mathrm{S}$, $N_\mathrm{R}$ transmit, scatter, and receive antennas, respectively.
	This configuration is denoted as $N_\mathrm{T} \times N_\mathrm{S} \times N_\mathrm{R}$ in the following context.
	The \gls{bd}-\gls{ris} can be modeled as an $N_\mathrm{S}$-port network \cite{Ivrlac2010} that further divides into $G$ individual groups, each containing $L \triangleq N_\mathrm{S} / G$ elements interconnected by real-time reconfigurable components \cite{Shen2020a}.
	To simplify the analysis and explore the performance limits, we assume a lossless asymmetric network without mutual coupling between scattering elements, as previously considered in \cite{Li2023b,Li2023c,Bartoli2023}.
	The overall scattering matrix of the \gls{bd}-\gls{ris} is block-unitary
	\begin{equation}
		\mathbf{\Theta} = \mathrm{diag}(\mathbf{\Theta}_1,\ldots,\mathbf{\Theta}_G),
		\label{eq:bd_ris}
	\end{equation}
	where $\mathbf{\Theta}_g \in \mathbb{U}^{L \times L}$ is the $g$-th unitary block (i.e., $\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g = \mathbf{I}$) that describes the response of group $g \in \mathcal{G} \triangleq \{1, \ldots, G\}$.
	Note that diagonal and unitary \gls{ris} can be regarded as its extreme cases with group size $L=1$ and $L=N_\mathrm{S}$, respectively.
	Some potential physical architectures of \gls{bd}-\gls{ris} are illustrated in \cite[Fig. 3]{Shen2020a}, \cite[Fig. 5]{Li2023c}, and \cite[Fig. 2]{Nerini2024}, where the
	% radiation pattern and
	circuit topology need to be modelled in the scattering matrix.

@@

	% \begin{remark}
	% 	\gls{bd}-\gls{ris} reduces to diagonal \gls{ris} and unitary \gls{ris} with group size $L$ 1 and $N_\mathrm{S}$, respectively.
	% \end{remark}
	% \begin{remark}
	% 	Individual forward and backward \gls{csi} are required for \gls{bd}-\gls{ris} designs.
	% 	This is different from diagonal \gls{ris} where estimating their product is usually sufficient.
	% 	% Later we will show the potential benefits from the \gls{csi} overhead.
	% \end{remark}
\end{section}

\begin{section}{Group-Wise Geodesic \glsfmtshort{rcg}}
	In this section, we first provide an overview on signal processing techniques for general \gls{bd}-\gls{ris} design problems, then propose a novel group-wise geodesic \gls{rcg} method that exploits the properties of unitary group to operate directly on the Stiefel manifold.
	% The proposed method is applicable to any smooth optimization problem on the Stiefel manifold and is particularly suitable for \gls{bd}-\gls{ris} design problems.
	We will later show that the proposed method not only provides better objective value and faster convergence than other approaches, but also works for arbitrary group size and any smooth optimization problem.
	\begin{subsection}{Conventional Techniques}
		General \gls{bd}-\gls{ris} design techniques can be classified into two categories based on the optimization variable and hardware architecture:
		\begin{itemize}
			\item \emph{Scattering matrix $\mathbf{\Theta}$:} This approach is often exploited for asymmetric architectures where the feasible domain of each group is a $L$-dimensional Stiefel manifold $\mathbf{\Theta}_g \in \mathbb{U}^{L \times L}$. Due to its non-convexity, relevant problems are usually solved by general non-geodesic manifold optimization \cite{Li2023b,Li2023c,Bartoli2023} or relax-then-project methods \cite{Fang2023}. The former will be discussed in the next subsection.
			\item \emph{Reactance matrix $\mathbf{X}$:\footnote{Similar design over susceptance matrix $\mathbf{B}$ can be found in \cite{Nerini2024}.}} This approach is often exploited for symmetric architectures where every pair of elements in the same group are connected by capacitors and inductors. According to network theory \cite{Pozar2011}, it maps to the scattering matrix by $\mathbf{\Theta}_g = (\jmath \mathbf{X}_g + Z_0 \mathbf{I})^{-1} (\jmath \mathbf{X}_g - Z_0 \mathbf{I})$, which formulates an unconstrained optimization problem on the upper triangular entries of $\{\mathbf{X}_g\}_{g \in \mathcal{G}}$ that is solvable by quasi-Newton methods \cite{Shen2020a}.
		\end{itemize}
	\end{subsection}

	\begin{subsection}{Geodesic vs Non-Geodesic \glsfmtshort{rcg}}
		We first revisit the general non-geodesic \gls{rcg} method that is applicable to optimization problems over arbitrary manifolds \cite{Absil2009,Pan2022d}.

>>>>

\usepackage[table]{xcolor}
\usepackage{adjustbox}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bookmark}
\usepackage{booktabs}
\usepackage[makeroom]{cancel}
\usepackage[american]{circuitikz}
\usepackage{cite}
% \usepackage{fixmath}
\usepackage[acronym]{glossaries-extra}
\usepackage{hyperref}
\usepackage{import}
\usepackage{mathtools}
\usepackage{microtype}
\usepackage[short]{optidef}
\usepackage{pgfplots}
\usepackage{ragged2e}
\usepackage[subtle]{savetrees}
\usepackage{siunitx}
\usepackage{stfloats}
\usepackage[caption=false,font=footnotesize,subrefformat=parens,labelformat=parens]{subfig}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{multirow}
\usepackage{booktabs}
% \usepackage{cleveref}

@@

\begin{document}
\title{Channel Shaping Using Beyond Diagonal Reconfigurable Intelligent Surfaces}
\author{
	\IEEEauthorblockN{
		Yang~Zhao,~\IEEEmembership{Member,~IEEE,}
		Hongyu~Li,~\IEEEmembership{Graduate Student Member,~IEEE,}\\
		Massimo~Franceschetti,~\IEEEmembership{Fellow,~IEEE,}
		and~Bruno~Clerckx,~\IEEEmembership{Fellow,~IEEE}
	}
	\thanks{
		Yang Zhao, Hongyu Li, and Bruno Clerckx are with the Department of Electrical and Electronic Engineering, Imperial College London, London SW7 2AZ, U.K. (e-mail: \{\href{mailto:yang.zhao18@imperial.ac.uk}{yang.zhao18}, \href{mailto:c.li21@imperial.ac.uk}{c.li21}, \href{mailto:b.clerckx@imperial.ac.uk}{b.clerckx}\}@imperial.ac.uk).

@@

\begin{section}{Introduction}
	Today we are witnessing a paradigm shift from connectivity to intelligence, where the wireless environment is no longer a chaotic medium but a conscious agent that can serve on demand.
	This is empowered by the recent advances in \gls{ris}, a programmable passive planar surface that recycles and redistributes ambient electromagnetic waves for improved wireless performance.
	A typical \gls{ris} consists of numerous low-power sub-wavelength non-resonant scattering elements, whose response can be engineered in real-time to manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019}.
	% It not only experiences negligible noise and supports full-duplex transmissions, but also features better flexibility than reflectarrays, lighter footprint than various relays, and greater scalability than conventional multi-antenna techniques.
	It enables low-noise, full-duplex operation, and also features better flexibility than reflectarrays, lighter footprint than various relays, and greater scalability than conventional \gls{mimo} systems.
	The most popular \gls{ris} research direction is \emph{joint passive and active beamforming} design with transceivers for a specific performance measure, which has attracted significant attention in wireless communication \cite{Wu2019,Guo2020,Liu2022}, sensing \cite{He2022,Luo2022,Hua2023}, and power transfer literature \cite{Wu2020a,Feng2022,Zhao2022}.
	While passive beamforming at \gls{ris} suffers severe attenuation from double fading, it offers better asymptotic behaviors than active beamforming at transceivers (e.g., second-order array gain and fourth-order harvested power \cite{Zhao2022}).
	Another feature is that \gls{ris} can also be used for \emph{backscatter modulation} by periodically switching its reflection pattern within the channel coherence time.
	This creates a free-ride message stream (similar to index modulation \cite{Basar2017}) with dual benefits: integrating with the legacy transmitter for enhanced channel capacity \cite{Karasik2020,Basar2020,Ye2022} or serving as a dedicated source for low-power uplink communication \cite{Liang2020,Zhao2024,Yang2024}.
	Different from above directions, \emph{channel shaping} exploits the \gls{ris} as a stand-alone device to modify the inherent properties of the wireless environment.
	To name a few, it can compensate for the Doppler effect \cite{Basar2021}, transform frequency-selective channels into frequency-flat \cite{Arslan2022}, improve the spatial diversity for \gls{mimo} systems \cite{Ozdogan2020a}, and create artificial time diversity for orthogonal \cite{Yang2019} and non-orthogonal \cite{Chen2023} multiple accesses.
	Channel shaping also provides a ubiquitous benchmark for different wireless applications and helps to decouple the \gls{ris}-transceiver design in a shape-then-transmit manner.
	At a specific time and frequency, channel metrics can be roughly classified into two categories:
	% shaping metrics can be classified into two categories:
	\begin{itemize}
		\item \emph{Singular value:} Relevant metrics are in general closely related to the performance measures (e.g., achievable rate and harvested power \cite{Shen2021}) but sensitive to numerical perturbations of the channel matrix. The impact of \gls{ris} has been studied in terms of minimum singular value \cite{ElMossallamy2021}, effective rank \cite{ElMossallamy2021,Meng2023}, condition number \cite{Zheng2022,Huang2023}, and degree of freedom \cite{Bafghi2022,Zheng2023,Chae2023}.
		\item \emph{Power:} Second-order statistics are less informative in \gls{mimo} but easier to analyze and optimize. The impact of \gls{ris} has been studied in terms of channel power gain \cite{Wu2019,Shen2020a,Nerini2023,Nerini2024,Santamaria2023} and leakage interference \cite{Santamaria2023a}.
	\end{itemize}

	Although those works offer initial glimpses into the channel shaping potential, one critical question has not been fully addressed: \emph{What is the ultimate channel shaping capability of a passive \gls{ris} in terms of singular values?}

@@

	Most relevant works \cite{ElMossallamy2021,Meng2023,Zheng2022,Huang2023,Bafghi2022,Zheng2023,Chae2023,Wu2019,Santamaria2023a} considered a diagonal \gls{ris} model where each element is connected to a dedicated impedance and is disconnected from others, such that wave impinging on one element is entirely reflected by the same element.
	% is tuned by a dedicated impedance and acts as an \emph{individual} scatterer \cite{Wu2020}.
	This simple architecture mathematically translates to a diagonal scattering matrix with unit-magnitude entries on the main diagonal and zeros elsewhere, applying merely a phase shift to the incoming signal.
	The concept was soon generalized to \gls{bd}-\gls{ris} with a group-connected architecture \cite{Shen2020a} that groups adjacent elements and connects them via passive reconfigurable components.\footnote{Those components can be either symmetric (e.g., capacitors and inductors) or asymmetric (e.g., ring hybrids and branch-line hybrids) \cite{Ahn2006}, resulting in symmetric and asymmetric scattering matrices, respectively.\label{fn:symmetric}}
	This allows wave impinging on one element to propagate within the circuit and depart partially from any element in the same group.

@@

	The authors of \cite{Bartoli2023} investigated the rate-optimal joint beamforming design for a specific \gls{bd}-\gls{ris}-aided \gls{mimo} system with blocked direct link and fully-connected \gls{ris}.

@@

	Moreover, no previous work has attempted to characterize the singular value region of a \gls{mimo} channel manipulated by any type of \gls{ris}, and the \gls{bd}-\gls{ris}-aided \gls{mimo} rate maximization problem remains unsolved.

@@

	First, we interpret the channel shaping potential of \gls{bd}-\gls{ris} in terms of channel rearrangement and space alignment.
	Channel rearrangement refers to rearranging and recombining the forward and backward channel branches (i.e., entries of the channel matrix) according to their strength.
	Space alignment generalizes phase matching in \gls{siso} and \gls{miso} to the high-dimensional singular vector space in \gls{mimo}.
	The off-diagonal entries of \gls{bd}-\gls{ris} scattering matrix provides a higher design freedom and exploits channel spatial diversity more effectively than diagonal \gls{ris}.
	This is the first paper to study \gls{bd}-\gls{ris} in general \gls{mimo} systems.
	%  and discuss its shaping potential.

	Second, we propose an efficient \gls{bd}-\gls{ris} design framework that solves general optimization problems by geodesic\footnote{A geodesic refers to the shortest path between two points in a Riemannian manifold.} \gls{rcg}.
	% to numerically analyze the channel shaping capability of \gls{bd}-\gls{ris}, we exploit the Riemannian geometry of the Stiefel manifold and
	This method modified from \cite{Abrudan2008,Abrudan2009} exploits the Riemannian geometry of the Stiefel manifold and performs group-wise multiplicative rotational updates along the geodesics, which avoids retraction and facilitates step size selection.
	It thus converges faster than existing non-geodesic approaches \cite{Absil2009,Pan2022d}.
	This is the first work to tailor an efficient optimization framework for \gls{bd}-\gls{ris}.

	% pioneer
	Third, we adopt the proposed geodesic \gls{rcg} and provide a numerical answer to the singular value shaping question.
	% to answer the channel singular value shaping question via numerical optimization.
	% to redistribute the singular values of a point-to-point \gls{mimo} channel.
	The {Pareto frontiers} of channel singular values are characterized by solving a series of weighted sum maximization problems, where the weight associated with each singular value can be positive, zero, or negative.
	% optimizing their weighted sum, where the weights can be positive, zero, or negative.
	% This problem is solved by the proposed geodesic \gls{rcg} algorithm.
	The enclosed region generalizes most relevant metrics and provides an intuitive channel shaping benchmark.
	Results show that increasing \gls{bd}-\gls{ris} group size enlarges the singular value region, improving both dynamic range of and trade-off between singular values.
	% , where the dynamic range and trade-off between are both improved.
	% significantly enlarges the singular value region, improveing the dynamic range and trade-off in-between.
	% We also and derive some analytical singular value bounds for rank-deficient \gls{mimo} and fully-connected \gls{ris}.

	Fourth, we leverage matrix analysis tools and provide an analytical answer to the singular value shaping question.
	% revisit the channel singular value shaping problem from an analytical perspective via matrix analysis.
	% To explore the ultimate shaping limits, we derive individual and collective singular value bounds for rank-deficient forward/backward channel and blocked direct channel.
	% Asymptotic bounds shared by diagonal and \gls{bd}-\gls{ris} are derived in the former case, while exact bounds for unitary \gls{bd}-\gls{ris} are derived in the latter case.
	Asymptotic channel singular value bounds shared by diagonal and \gls{bd}-\gls{ris} are derived for rank-deficient forward/backward channel.
	On the other hand, attainable bounds in the presence of fully-connected \gls{ris} are derived for blocked direct channel.
	Results validate those bounds and emphasizes the shaping advantage of \gls{bd}-\gls{ris}.
	This is the first work to quantify the singular value redistribution capability from either numerical or analytical perspective.
	% Results validate those bounds and show that
	% fully-connected \gls{bd}-\gls{ris} without direct link.

@@

	% % respectively in rank-deficient channels and for fully-connected \gls{bd}-\gls{ris}
	% collective singular value bounds	for rank-deficient channels and fully-connected \gls{ris}, respectively.
	% 	Results validate those bounds and show
	% for rank-deficient channels and fully-connected \gls{ris}, respectively.
	% Results validate those bounds and show that increasing \gls{bd}-\gls{ris} group size significantly enlarges the singular value region, providing wider dynamic range and better trade-off.
	% dynamic range and trade-off of singular values are significantly improved by \gls{bd}-\gls{ris}, especially in large-scale \gls{mimo} systems.
	% This is the first work to comprehensively answer the channel shaping capability question from both numerical and analytical perspectives in the most general \gls{bd}-\gls{ris} setup.
	% from a Pareto perspective.

	Fifth, we tackle \gls{bd}-\gls{ris}-aided \gls{mimo} rate maximization problem by two beamforming solutions: a local-optimal \gls{ao} approach and a low-complexity approach inspired by channel shaping insights.
	The former updates active beamforming by eigenmode transmission and passive beamforming by geodesic \gls{rcg} until convergence.
	The latter suboptimally decouples the problem as channel power gain maximization at \gls{ris} and conventional \gls{mimo} transmission at transceiver, then solves both in closed form.
	% The problem is solved by a closed-form two-stage method.
	Results show that \gls{bd}-\gls{ris} provides higher power gain and achievable rate than diagonal \gls{ris}, and the rate gap between two approaches diminishes as the \gls{ris} evolves from single-connected to fully-connected.
	It suggests channel shaping offers a promising path to decouple joint \gls{ris}-transceiver designs.
	% Interestingly, the rate difference between the optimal and low-complexity designs diminishes as the \gls{ris} evolves from diagonal to unitary.
	% Those results emphasize the importance of using \gls{bd}-\gls{ris} in large-scale \gls{mimo} systems and suggest channel shaping offers a promising path to decouple joint \gls{ris}-transceiver designs.
	% a promising channel shaping path towards practical \gls{ris} and transceiver design.
	% that channel shaping offers a promising path towards simplified and practical \gls{bd}-
	% It suggests that channel shaping offers a promising path towards simplified and practical \gls{bd}-\gls{ris} (and transceiver) designs.
	% via \gls{bd}-\gls{ris} offers a promising path towards simplified and practical design.
	% offers a promising path towards simplified and practical desi
	% It suggests channel shaping helps to decouple the joint \gls{ris}-transceiver design and offers a promising low-complexity solution.
	% This is the first work to discuss the implications of channel shaping on joint active-passive beamforming optimization.

@@

\begin{section}{\glsfmtshort{bd}-\glsfmtshort{ris} Model}
	Consider a \gls{bd}-\gls{ris} aided point-to-point \gls{mimo} system with $N_\mathrm{T}$, $N_\mathrm{S}$, $N_\mathrm{R}$ transmit, scatter, and receive antennas, respectively.
	This configuration is denoted as $N_\mathrm{T} \times N_\mathrm{S} \times N_\mathrm{R}$ in the following context.
	The \gls{bd}-\gls{ris} can be modeled as an $N_\mathrm{S}$-port network \cite{Ivrlac2010} that further divides into $G$ individual groups, each containing $L \triangleq N_\mathrm{S} / G$ elements interconnected by real-time reconfigurable components \cite{Shen2020a}.
	To simplify the analysis and explore the performance limits, we assume a lossless asymmetric network without mutual coupling between scattering elements, as previously considered in \cite{Li2023b,Li2023c,Bartoli2023}.
	The overall scattering matrix of the \gls{bd}-\gls{ris} is block-unitary\footnote{Following footnote \ref{fn:symmetric}, we do not assume the scattering matrix to be symmetric. If required, one can enforce symmetry over the block-unitary result by $\mathbf{\Theta} \gets (\mathbf{\Theta} + \mathbf{\Theta}^\mathsf{T})/2$.}
	\begin{equation}
		\mathbf{\Theta} = \mathrm{diag}(\mathbf{\Theta}_1,\ldots,\mathbf{\Theta}_G),
		\label{eq:bd_ris}
	\end{equation}
	where $\mathbf{\Theta}_g \in \mathbb{U}^{L \times L}$ is the $g$-th unitary block (i.e., $\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g = \mathbf{I}$) that describes the response of group $g \in \mathcal{G} \triangleq \{1, \ldots, G\}$.
	Note that diagonal \gls{ris} can be regarded as its extreme case with group size $L=1$.
	Some potential physical architectures of \gls{bd}-\gls{ris} are illustrated in \cite[Fig. 3]{Shen2020a}, \cite[Fig. 5]{Li2023c}, and \cite[Fig. 2]{Nerini2024}, where the
	% radiation pattern and
	circuit topology need to be modelled in the scattering matrix.

@@

	% \begin{remark}
	% 	\gls{bd}-\gls{ris} reduces to diagonal \gls{ris} and fully-connected \gls{ris} with group size $L$ 1 and $N_\mathrm{S}$, respectively.
	% \end{remark}
	% \begin{remark}
	% 	Individual forward and backward \gls{csi} are required for \gls{bd}-\gls{ris} designs.
	% 	This is different from diagonal \gls{ris} where estimating their product is usually sufficient.
	% 	% Later we will show the potential benefits from the \gls{csi} overhead.
	% \end{remark}
\end{section}

\begin{section}{\glsfmtshort{bd}-\glsfmtshort{ris} Design by Geodesic \glsfmtshort{rcg}}
	% In this section, we first provide an overview on signal processing techniques for group-connected \gls{bd}-\gls{ris} design problems, then propose a novel group-wise geodesic \gls{rcg} method that exploits the properties of unitary group and operates directly on the Stiefel manifold.
	% The proposed method is applicable to any smooth optimization problem on the Stiefel manifold and is particularly suitable for \gls{bd}-\gls{ris} design problems.
	% We will later show that the proposed method provides better objective value and faster convergence for smooth optimization problems of arbitrary group size.
	A geodesic is a curve representing the shortest path between two points in a Riemannian manifold, whose tangent vectors remain parallel when transporting along the curve.
	We notice that the feasible domain of scattering matrix of each \gls{bd}-\gls{ris} group is a $L$-dimensional Stiefel manifold $\mathbf{\Theta}_g \in \mathbb{U}^{L \times L}$, which is non-convex and non-Euclidean.
	Therefore, relevant optimization problems are usually solved by relax-then-project methods \cite{Fang2023} or universal non-geodesic manifold \gls{rcg} \cite{Li2023b,Li2023c,Bartoli2023}.
	The former solves an unconstrained version of original problem by quasi-Newton methods, then projects the solution back to the Stiefel manifold.
	It has no guarantee of optimality and may suffer from numerical instability.
	The latter generalizes the conjugate gradient methods to any Riemannian manifold, which updates the feasible point by progressing along the conjugate direction and projecting back to the manifold at each step.
	In this section, we first discuss the existing non-geodesic \gls{rcg} method and its drawbacks, then propose a novel group-wise geodesic \gls{rcg} method that operates directly on the Stiefel manifold for faster convergence.

	\begin{subsection}{Non-Geodesic \glsfmtshort{rcg}}
		This universal \gls{rcg} method is applicable to optimization problems over arbitrary manifolds \cite{Absil2009,Pan2022d}.
